---
title: "session09 - exercises"
format: html
editor: visual
date: 2026-03-22
author:
  - Timo Roettger
  - Hågard Meyer  
execute:
  error: false
  warning: false
  message: false
  cache: false
bibliography: ../../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

# Introduction

The main **learning goals** of this week's practical parts are:

-   getting an intuition about the gap between verbal hypotheses and formal hypotheses

# The data

In this dataset, we will be analyzing data from @kilpatrick2023random:

In a nutshell, the authors were interested in whether there is a relationship between the form of a word and its meaning by looking at Japanese Pokemon names.

Let's have a look at the data:

```{r}
#| label: data

# Load in the dataset and already remove irrelevant stuff:
pokemon <- read_csv('../../data/Japanese_Pokemon_simplified.csv') |> 
  select(-`Sum of values`, -Checksum)

# check data set:
pokemon

```

What you see here are the following variables:

Variables:

`Designation`: source of Pokemon name. Official = the official names of Japanese Pokemon; Elicited = elicited names from naive participants.

(`pokemonname`: Pokemon name in Japanese) removed in simplified data file

`Output`: Pokemon name in transcribed form

`Length`: Amount of phones in name

`Evolution`: Evolution stage of Pokemons: 1 = pre-evolution; 2 = post-evolution, NA = do not evolve

`f` = amount of /f/ sounds in name

`T` = amount of /T/ sounds in name

...

and many more different sounds.

Now imagine researchers want to use this data set to *test* the following hypothesis:

*The evolution stage of a Pokemon is reflected in aspects of its Japanese name*

# Exercise

## A: Preregister your analysis

Take the hypothesis and what you know about the data set and its structure to predict a very specific relationship that you can test. You are allowed to look at the dataset, but do not look at concrete relationships that might be a manifestation of the hypothesis.

### My approach

Taking the hypothesis given above I instinctively see that it is vague. One aspect I see that is testable is the relationship between evolution stage and length of Japanese name. This can be done simply by predicting the length of Japanese name by the evolution stage, including a random effect for the whether the output is from the Pokemon official name, or an elicited one. Pokemon that cannot evolve are removed. An addition of random effects for both unique Pokemon and speaker for elicited names could have been considered, but, as an equal output does not ensure that it is the same Pokemon, this is not done due to there not being a variable for this.

The task is provided with the LME4 package, so I am assuming we are doing the comparison using this.

## B: Run your preregistered analysis

Run the preregistered analysis made in (A) above and interpret your results relative to the hypothesis.

### My approach

```{r}

# Removing pokemon that cannot evolve and turning the designation into a factor
pokemon_clean <- 
  pokemon %>% 
  filter( !is.na ( Evolution ) ) %>% 
  mutate ( Designation = as.factor ( Designation ) )

# Fitting a model to the relationship
xmdl <- lmer ( Length ~ 
               Evolution + 
               (1 | Designation ),
             data = pokemon_clean)

# Fitting a null model to compare it to
xmdl_null <- lmer ( Length ~ 
                      1 + 
                      (1 | Designation ),
                    data = pokemon_clean)

# Looking at the model
summary(xmdl)

# Testing the model against a null hypothesis
anova(xmdl_null, xmdl)
```

Looking at the output we see that by moving up a stage in evolution we get an increase of about \~1.27 in name length. Mean length of pre-evolution Pokemon is predicted to be \~6.99 ( \~5.72 + \~1.27 ) and for post-evolution Pokemon we get \~8.27 ( \~5.72 + ( 2 \* \~1.27 ) ). A P-value of less than 0.001 gives some credibility to this claim.

## C: Show contradicting results

Now explore the data and find two alternative way to test the research hypothesis that yield convincing results that go in opposite directions based of your findings in (A). Justify all your choices verbally after the fact.

### My approach

Exploring the dataset I find some concerning details (in addition to the lack of specifier noted in A). As an example, I can find an output "rizaRdo" noted with NA in Evolution to mark "do not evolve". I suspect this is リザード (Hepburn "rizādo") who does evolve into リザードン (Hepburn "rizādon"). Not having a species definition in the data set, I have to analyse what I am given. I choose to look at the relationship between Evolution and 1. the proportion of unique sounds in the Japanese name 2. the proportion of consonants in the Japanese name.

```{r}

# Additional mutations

pokemon_mutated <- 
  pokemon_clean %>% 
  mutate ( unique_characters = rowSums ( select ( .,
                                                  f:c )
                                         > 0 ) ) %>% 
  mutate ( proportion_unique_characters = unique_characters / Length ) %>% 
  mutate ( sum_vowels = rowSums ( select ( .,
                                           a,
                                           i,
                                           u,
                                           e,
                                           o,
                                           R) ) ) %>% 
  mutate ( proportion_vowels = sum_vowels / Length ) %>% 
  mutate ( proportion_consonants = 1 - proportion_vowels )

# Fitting a model to predict proportion of unique characters based on Evolution
ymdl <- lmer ( prop_unique ~ 
                 Evolution + 
                 (1 | Designation ),
               data = pokemon_mutated)

# Fittin null model
ymdl_null <- lmer ( prop_unique ~ 
                      1 + 
                      (1 | Designation ),
                    data = pokemon_mutated)

# Showing model
summary(ymdl)

# Comparing model to null
anova(ymdl_null, ymdl)

# Fitting model to predict proportion of consonants based on Evolution
zmdl <- lmer ( proportion_consonants ~ 
                 Evolution + 
                 (1 | Designation ),
               data = pokemon_mutated)

# Fitting null model
zmdl_null <- lmer ( proportion_consonants ~ 
                      1 + 
                      (1 | Designation ),
                    data = pokemon_mutated)

# Showing model
summary(zmdl)

# Comparing model to null
anova(zmdl_null, zmdl)
```

My justification for choosing to look at the proportion of unique characters is that this can be interpreted as an increase in complexity in the Japanese name. If evolution marks an increase in features and / or power, we might expect an increase in complexity or diversity to reflect that. The model does, however, provide a negative trend for proportion of unique character as a Pokemon evolves ( -\~0.029 with a P-value of less than 0.001). This indicates that Pokemon that can evolve have more variation in characters than post-evolution Pokemon.

My justification for choosing to look at the proportion of consonants in the Japanese name is that consonants can be perceived as harsher and / or stronger than vowels. This can be thought to mirror the pre- and post-evolution Pokemon change given the presumed change in power. Again the model predicts a negative trend here ( \~-0.010 with a P-value of \~0.008) suggesting that there is a lower proportion of consonants in the post-evolution.

## D: Researcher degrees of freedom

Note down all dimensions of researcher degrees of freedom that make the mapping of the hypothesis onto your data analysis vague. How many possible combinations can you think of?

### My approach

I will do this by taking the hypothesis as sections and judging those.

> *The evolution stage of a Pokemon is reflected in aspects of its Japanese name*

*The evolution stage* is vague as is. Mainly, what happens with Pokemon that cannot evolve? These can be interpreted as irrelevant (not part of an evolution stage), possibly joined with pre-evolution Pokemon (they have not evolved) or joined with the post-evolution Pokemon (they cannot evolve further). Looking at the Evolution column it gets worse. I have mentioned that some Pokemon seem to show wrong information about their ability to evolve. Furthermore, prior knowledge of Pokemon tells me that Pokemon can have more categories than are present in the dataset. at the very least we have:

1.  Pokemon that do not evolve and have not evolved from anything
2.  Pokemon that can evolve and have not evolved from anything
3.  Pokemon that can evolve and have evolved from something
4.  Pokemon that cannot evolve and have evolved from something

*Pokemon* also lends itself to some vagueness here. Looking at the data set we have two different designations for these, namely *Official* (presumably a written official name) and *Elicited* (as by a naive speaker). This ties together with the wording *Japanese name*. A researcher can either join these as one group, or choose to focus on just one of them. While these are likely connected, they are not guaranteed to give the same outputs.

Finally *aspect* is likely the most vague term. A name can have an almost infinite number of aspects, like intonation, stress, connotations to other words etc. Even within the given data set, a researcher can choose to look at length of name, any one of the output sounds individually or any number of combination and orderings of these. The number of combinations here is still astronomical and each of those combinations can be combined again with decisions made in relation to my previous paragraphs decisions.

Importantly these choices are not trivial. While they might lead to similar conclusions, they just as well might not.

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
