---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

# Contrast coding

contrasts(sim$Sem) <- contr.sum(2)
contrasts(sim$Phon) <- contr.sum(2)

# Square root of distance to combat slight model mismatch
sim <- sim %>% 
  mutate ( Distance_sqrt = sqrt ( Distance ) )

# Formula without random effects as they are not avaliable in the dataset
my_formula <- bf(Distance_sqrt ~ Sem * Phon)

# Calculating weak priors with the knowledge that the task was to mark two points on an
# A4 sheet of paper, and measured in mm
a <- 297
b <- 210
iterations <- 1000000
set.seed(333)

# Generate poin-pairs on the sheet
x1 <- runif(iterations, 0, a)
x2 <- runif(iterations, 0, b)
y1 <- runif(iterations, 0, a)
y2 <- runif(iterations, 0, b)

# Distance between two points
distance <- sqrt(
    (x2 - x1)^2
    +
    (y2 - y1)^2)

mean(distance)
sd(distance)

# Longest distance on A4
sqrt ( a^2 + b^2 )


get_prior(my_formula, sim)

# Now priors for sqrt of Distance
my_priors <- c( prior(normal(12,3), class = "Intercept"),
                prior(normal(0,3), class = "b"),
                # Sigma gets 2*SD as I do not want it too restrictive
                prior(normal(0,6), class = "sigma")
              )


# Testing prior
xmdl_prior <- brm ( formula = my_formula,
                    prior = my_priors,
                    sample_prior = "only",
                    data = sim,
                    seed = 333)

# Visualizing my priors

xmdl_prior %>%
  add_epred_draws(newdata = expand.grid(Sem  = levels(sim$Sem),
                                        Phon = levels(sim$Phon))) %>%
  ggplot(aes(x = .epred, 
             color = interaction(Sem, Phon))) +
  geom_density(size = 1.3) +
  labs(title = "Prior predicted distances by condition",
       x     = "Square root of distance in mm",
       y     = "Density",
       color  = NULL) +
  scale_color_manual(values = c("Different.Different" = '#00FFFF',
                               "Similar.Different" = '#FF00FF',
                               "Different.Similar" = '#FFFF00',
                               "Similar.Similar" = '#000000'),
                    labels=c("Sem:Diff Phon:Diff",
                             "Sem:Sim Phon:Diff",
                             "Sem:Diff Phon:Sim",
                             "Sem:Sim Phon:Sim")) + 
  theme_minimal() +
  theme(legend.position = "bottom")

```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

xmdl <- brm ( formula = my_formula,
              family  = gaussian(),
              data    = sim,
              chains  = 4,
              iter    = 2000,
              warmup  = 1000,
              seed    = 333
            )


```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

#YOUR CODE HERE
pp_check(xmdl, ndraws = 200)

# Going back to sqrt the Distance
```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# Finding caterpillars and RHat / ESS. RHat of 1.00 and smallest ESS = 2530
mcmc_plot(xmdl, type = "trace")
summary(xmdl)
```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

# Plotting posterior
xmdl %>%
  add_epred_draws(newdata = expand.grid(Sem  = levels(sim$Sem),
                                        Phon = levels(sim$Phon))) %>%
  ggplot(aes(x = .epred, 
             color = interaction(Sem, Phon))) +
  geom_density(size = 1.3) +
  geom_density(
    data = xmdl_prior %>%
      add_epred_draws(newdata = expand.grid(Sem = levels(sim$Sem),
                                            Phon = levels(sim$Phon))),
    aes(x = .epred, color = interaction(Sem, Phon)),
    linetype = "dashed",
    size = 1
  ) +
  coord_cartesian(xlim = c(7,10)) + 
  labs(title = "Posterior predicted distances by condition",
       subtitle = "Prior prediction as dashed lines"
       x     = "Square root of distance in mm",
       y     = "Density",
       color  = NULL) +
  scale_color_manual(values = c("Different.Different" = '#00FFFF',
                               "Similar.Different" = '#FF00FF',
                               "Different.Similar" = '#FFFF00',
                               "Similar.Similar" = '#000000'),
                    labels=c("Sem:Diff Phon:Diff",
                             "Sem:Sim Phon:Diff",
                             "Sem:Diff Phon:Sim",
                             "Sem:Sim Phon:Sim")) + 
  theme_minimal() +
  theme(legend.position = "bottom")


```

## (g) Write up your entire analysis:

YOUR WRITE UP HERE

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
