---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

# Contrast coding

contrasts(sim$Sem) <- contr.sum(2)
contrasts(sim$Phon) <- contr.sum(2)

# Square root of distance to combat slight model mismatch
sim <- sim %>% 
  mutate ( Distance_sqrt = sqrt ( Distance ) )

# Formula without random effects as they are not avaliable in the dataset
my_formula <- bf(Distance_sqrt ~ Sem * Phon)

# Calculating weak priors with the knowledge that the task was to mark two points on an
# A4 sheet of paper, and measured in mm
a <- 297
b <- 210
iterations <- 1000000
set.seed(333)

# Generate poin-pairs on the sheet
x1 <- runif(iterations, 0, a)
x2 <- runif(iterations, 0, b)
y1 <- runif(iterations, 0, a)
y2 <- runif(iterations, 0, b)

# Distance between two points
distance <- sqrt(
    (x2 - x1)^2
    +
    (y2 - y1)^2)

mean(distance)
sd(distance)

# Longest distance on A4
sqrt ( a^2 + b^2 )


get_prior(my_formula, sim)

# Now priors for sqrt of Distance
my_priors <- c( prior(normal(12,3), class = "Intercept"),
                prior(normal(0,3), class = "b"),
                # Sigma gets 2*SD as I do not want it too restrictive
                prior(normal(0,6), class = "sigma")
              )


# Testing prior
xmdl_prior <- brm ( formula = my_formula,
                    prior = my_priors,
                    sample_prior = "only",
                    data = sim,
                    seed = 333)

# Visualizing my priors

xmdl_prior %>%
  add_epred_draws(newdata = expand.grid(Sem  = levels(sim$Sem),
                                        Phon = levels(sim$Phon))) %>%
  ggplot(aes(x = .epred, 
             color = interaction(Sem, Phon))) +
  geom_density(size = 1.3) +
  labs(title = "Prior predicted distances by condition",
       x     = "Square root of distance in mm",
       y     = "Density",
       color  = NULL) +
  scale_color_manual(values = c("Different.Different" = '#00FFFF',
                               "Similar.Different" = '#FF00FF',
                               "Different.Similar" = '#FFFF00',
                               "Similar.Similar" = '#000000'),
                    labels=c("Sem:Diff Phon:Diff",
                             "Sem:Sim Phon:Diff",
                             "Sem:Diff Phon:Sim",
                             "Sem:Sim Phon:Sim")) + 
  theme_minimal() +
  theme(legend.position = "bottom")

```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

xmdl <- brm ( formula = my_formula,
              family  = gaussian(),
              data    = sim,
              prior   = my_priors,
              chains  = 4,
              iter    = 2000,
              warmup  = 1000,
              seed    = 333
            )


```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

# Cheking fit
pp_check(xmdl, ndraws = 200)

```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# Finding caterpillars and RHat / ESS. RHat of 1.00 and smallest ESS = 2530
mcmc_plot(xmdl, type = "trace")
summary(xmdl)
```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

# Plotting posterior
xmdl %>%
  add_epred_draws(newdata = expand.grid(Sem  = levels(sim$Sem),
                                        Phon = levels(sim$Phon))) %>%
  ggplot(aes(x = .epred, 
             color = interaction(Sem, Phon))) +
  geom_density(size = 1.3) +
  geom_density(
    data = xmdl_prior %>%
      add_epred_draws(newdata = expand.grid(Sem = levels(sim$Sem),
                                            Phon = levels(sim$Phon))),
    aes(x = .epred, color = interaction(Sem, Phon)),
    linetype = "dashed",
    size = 1
  ) +
  coord_cartesian(xlim = c(7,10)) + 
  labs(title = "Posterior predicted distances by condition",
       subtitle = "Prior prediction as dashed lines",
       x     = "Square root of distance in mm",
       y     = "Density",
       color  = NULL) +
  scale_color_manual(values = c("Different.Different" = '#00FFFF',
                               "Similar.Different" = '#FF00FF',
                               "Different.Similar" = '#FFFF00',
                               "Similar.Similar" = '#000000'),
                    labels=c("Sem:Diff Phon:Diff",
                             "Sem:Sim Phon:Diff",
                             "Sem:Diff Phon:Sim",
                             "Sem:Sim Phon:Sim")) + 
  theme_minimal() +
  theme(legend.position = "bottom")

# Looking at posterior in numbers
posterior_summary(xmdl)

# Looking at direction

p_direction(xmdl)

# Calculating bayes factor
bayesfactor_parameters(xmdl)

```

## (g) Write up your entire analysis:

The goal is to test whether semantic similarity between cities, phonological 
similarity between cities and their interaction has an effect on the distance
between where participants draw the cities.

A Bayesian model was fitted predicting the square root of the distance
between cities from the phonological and semantic similarity, and their 
interaction. The choice to take the square root of the distance was made
to combat a slight right skew observed when running the model without.

I selected weakly informative priors based on the dimensions of a A4
sheet of paper and tested to show their predictions were within
a reasonable range.

The posterior predictive check was done visually and the predicted
data matched the observed data reasonably well. Sampling diagnostics
were also assessed visually with a trace plot that generated acceptable
caterpillars. Numerically the sampling produced no RHat above 1.00 and 
had a smallest effective sample size of 2500. This suggests that the
model is reliable.

Posteriors predict that the distance between cities tend to be smaller
when the names are semantically similar with a coefficient of 0.34. The 
credible intervals are 0.10 to 0.59. This is further corroborated by a 99.7%
probability that the direction is toward closed cities. Note, however, that
the Bayes factor for this is only 1.66, so care should be taken in making
too strong a claim for this hypothesis.

With a coefficient of -0.22 and credible intervals from -0.46 to 0.02 the 
effect of phonologically similar cities is much more uncertain. The predicted
direction is toward smaller distance, but with a probability of 96.67%. A
Bayes factor of 0.22 suggests weakly that there is no effect here.

The interaction is, while not clearly, more likely to be of no effect. With a 
coefficient of -0.05 (CI -0.29 to 0.20) and a probability of direction of 65.8%
this is close to zero. A Bayes factor of 0.046 means that there is a strong 
claim to be made for this not having an effect.

To conclude, the model suggests that there is some validity to the claim
that semantically similar cities are placed closer together, though not 
without some variance. Phonological similarity and the interaction between
the two are suggested through the model to have less of an impact, with some
uncertainty, though the model suggests that an effect of phonological similarity
is more likely than the effect from their interaction.

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
