---
title: "session09 - exercises"
format: html
editor: visual
date: 2026-03-18
author: 
  - Timo Roettger
  - Hågard Meyer
execute:
  error: false
  warning: false
  message: false
  cache: false
bibliography: ../../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, lme4, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

# Introduction

The main **learning goals** of this week's practical parts are:

-   getting an intuition about the gap between verbal hypotheses and formal hypotheses

# The data

In this dataset, we will be analyzing data from @kilpatrick2023random:

In a nutshell, the authors were interested in whether there is a relationship between the form of a word and its meaning by looking at Japanese Pokemon names.

Let's have a look at the data:

```{r}
#| label: data

# Load in the dataset and already remove irrelevant stuff:
pokemon <- read_csv('../../data/Japanese_Pokemon.csv') |> 
  select(-`Sum of values`, -Checksum)

# check data set:
pokemon

```

What you see here are the following variables:

Variables:

`Designation`: source of Pokemon name. Official = the official names of Japanese Pokemon; Elicited = elicited names from naive participants.

`pokemonname`: Pokemon name in Japanese

`Output`: Pokemon name in transcribed form

`Length`: Amount of phones in name

`Evolution`: Evolution stage of Pokemons: 1 = pre-evolution; 2 = post-evolution, NA = do not evolve

`f` = amount of /f/ sounds in name

`T` = amount of /T/ sounds in name

...

and many more different sounds.

Now imagine researchers want to use this data set to *test* the following hypothesis:

*The evolution stage of a Pokemon is reflected in aspects of its Japanese name*

# Exercise

## A: Preregister your analysis

Take the hypothesis and what you know about the data set and its structure to predict a very specific relationship that you can test. You are allowed to look at the dataset, but do not look at concrete relationships that might be a manifestation of the hypothesis.

### My approach

> *The evolution stage of a Pokemon is reflected in aspects of its Japanese name*

Taking this as a baseline I find it, likely by design, vague and in need of definition. Through prior knowledge I am aware that *Pokemon* are a set of fictional creatures, officially set by the *Pokedex* (@pokemon_zukan_2026), that has the ability to evolve into new forms. They also have names in Japanese. The problemetic vagueness comes with the usage of the term *aspect*, which can be interpreted and measured in a multitude of ways.

In order to make this a hypothesis I would consider acceptably researchable I have to choose an aspect and test its relationship. Prior knowledge of *Pokemon*, and evolution in general, leads me to view *evolution* as gaining a new feature. If these are to be reflected in the name, I suspect the names would have more unique sounds to show the increase in features. Based on this I can formulate a testable hypothesis.

> *Higher evolution Pokemon have a higher proportion of unique sounds in their Japanese name*

Moving to the data I get a

The number is suspiciously high however so I did a control against the official list of pokemon (@pokemon_zukan_2026).

```{r}

```

```{r}

# Adding a colum for the proportion of unique sounds in a name by taking 
# all columns that have a number bigger than zero and adding them together
# before deviding by the length column. This provides a proportion value
# bound on 1 (only unique characters) to the maximum length of the name
# (if the name only has a single character)
pokemon_mutated_nona <- pokemon %>%
  mutate ( proportion_unique_sounds = Length / rowSums ( select ( ., 
                                                                  f:c )
                                                         > 0 ) ) %>% 
  filter(!is.na(Evolution)) %>% 
  mutate(Evolution = as.factor(Evolution))

pokemon_mutated_nona
```

```{r}
my_formula <- bf(proportion_unique_sounds ~ Evolution)

get_prior(my_formula, pokemon_mutated_nona)

my_priors <- c( prior(normal(1.2,0.1), class = "Intercept"),
                prior(normal(0,0.1), class = "b"),
                prior(normal(0,2), class = "sigma")
              )

# Testing prior
xmdl_prior <- brm ( formula = my_formula,
                    prior = my_priors,
                    sample_prior = "only",
                    data = pokemon_mutated_nona,
                    seed = 333)

# Visualizing my priors

xmdl_prior %>%
  add_epred_draws(newdata = expand.grid(Evolution=c(1,2))) %>%
  ggplot(aes(x = .epred, 
             color = as.factor(Evolution) ) ) +
  geom_density(linewidth = 1.3) +
  labs(title = "AA",
       x     = "BB",
       y     = "CC",
       color  = NULL) +
  scale_color_manual(values = c("1" = '#FF00FF',
                               "2" = '#FFFF00'),
                    labels=c("Evo1",
                             "Evo2")) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

## B: Run your preregistered analysis

Run the preregistered analysis made in (A) above and interpret your results relative to the hypothesis.

## C: Show contradicting results

Now explore the data and find two alternative way to test the research hypothesis that yield convincing results that go in opposite directions based of your findings in (A). Justify all your choices verbally after the fact.

## D: Researcher degrees of freedom

Note down all dimensions of researcher degrees of freedom that make the mapping of the hypothesis onto your data analysis vague. How many possible combinations can you think of?

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
