---
title: "session04 - homework - solutions"
format: html
editor: visual
date: 2026-02-14
author: Hågard Meyer
execute:
  error: false
  warning: false
  message: false
  cache: false
bibliography: ../../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, ggeffects, patchwork)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

# Introduction to data

We will be working with perception data from one of my own experiments [@roettger2014assessing].

In this study, we investigated incomplete neutralization (IN). IN refers to phenomena in which a phonological neutralization is phonetically incomplete. In this paper, researchers looked at final devoicing in German. In German, the word "Räder" (*engl.* 'wheels') has a voiced /d/, but when forming the singular "Rad" (*engl.* 'wheel'), the /d/ is in word-final position and is devoiced, i.e. loses its phonological voicing and thus is supposed to be identical to the /t/ sound in "Rat" (*engl.* 'council'). However, when measuring the /d/ in "Rad" and the /t/ in "Rat", they are phonetically not identical. There are small but consistent differences between them. In experiment 4 of the paper, the researchers test whether these small acoustic differences can be perceived by naive listeners.

They showed 16 listeners 24 different pseudoword pairs produced by 16 different speakers and measured both reaction time as well as accuracy of correctly identifying the speaker-intended voicing category.

Let's load in the data and have a look:

```{r}
#| label: load-in

incomplete <- read_csv("../../data/incomplete_perception.csv") |> 
  # preprocessing the data such that everything that ought to be a factor is one
  mutate(listener = as.factor(listener),
         voice = as.factor(voice), 
         item_pair = as.factor(item_pair),
         correct_voicing = as.factor(correct_voicing),
         chosen_voicing = as.factor(chosen_voicing)
         )

# have a look
incomplete

```

These are the variables that you will find within the data:

-   `order`: trial order for each listener (1-392)
-   `listener`: unique identifier for each listener (n = 16)
-   `age`: listeners' age
-   `item_pair`: unique identifier for each word pair (e.g. pruk vs. prug, n = 24)
-   `voice`: unique identifier for each speaker voice (n = 16)
-   `RT` = reaction time in ms.
-   `ACC` = accuracy (binary, 1 = correct, 0 = incorrect)
-   `correct_voicing` = whether the audio stimuli contained an underlyingly voiced or voiceless stop in word-final position
-   `chosen_voicing` = the voicing category that the listeners selected in a 2-alternative forced-choice.

### (A) Run a mixed effects model

In their paper, they described different ways how to analyse the perception data. Among other things, they use a generalized linear mixed effects model. While not clearly articulated in the paper, they were mainly interested in whether listeners chose the intended voicing category more often than the unintended one. One way to test this hypothesis is to predict the `chosen_voicing` category by the `correct_voicing` category (i.e. the one that was intended by the speaker).

Build that model with the appropriate random effect structure, taking into account all important grouping variables.

(NOTE: if you encounter the following error message, ignore for now and rerun the model until it does not give this error: *Error in pwrssUpdate(pp, resp, tol = tolPwrss, GQmat = GQmat, compDev = compDev, : Downdated VtV is not positive definite*)

```{r}
#| label: glmer_incomplete


# Fitting a model to predict *chosen_voicing* based on *correct_voicing*
# Family is binomial because chosen_voicing is a factor with two features
xmdl <- glmer(
  chosen_voicing ~ 
    correct_voicing +
    (1 + correct_voicing | listener) +
    (1 + correct_voicing | item_pair) +
    (1 + correct_voicing | voice),
  data = incomplete,
  family = binomial
)

# Displaying a summary of the model
summary(xmdl)
```

### (B) Interpret the model output

Interpret the model coefficients and calculate p-values for the relevant predictor to test the hypothesis by comparing the full model (with the predictor) with the corresponding null model (without the predictor, i.e. just substitute the fixed effect predictor with 1).

```{r}
#| label: interpretation

# Calculating p-values by comparing it to the null model
# Making a null-model

xmdl_null <- glmer(
  chosen_voicing ~ 
    1 +
    (1 + correct_voicing | listener) +
    (1 + correct_voicing | item_pair) +
    (1 + correct_voicing | voice),
  data = incomplete,
  family = binomial
)

# Comparing the models with an ANOVA analysis
anova(xmdl_null, xmdl)
```

#### Interpretation

```{r}
#| label: probability_conversion

# Using plogis to calculate probability from logarithmic odds
plogis(c(
  -0.02359,           # (Intercept)
  0.52332            # correct_voicingvoiceless
))
```

The model predicts that the logarithmic odds of choosing (*chosen_voicing*) *voiced* when the stimuli (*correct_voicing*) was *voiced* was \~-0.024. Converting this to probability shows that participants chose *voiced* when the stimuli was *voiced* about 49% of the time. This is fairy close to a 50/50 chance split, which is also reflected in a p-value of \~0.934.

Further, it predicts that the logarithmic odds of choosing *voiced* when the stimuli was *voiceless* was changes  by \~0.523. Probability-wise this predicts this choice to happen about 63% of the time. This is likely significant (p-value of \~0.002), and implies that the category of stimuli affects the responses from participants. 

Comparing the model to the null-model shows that the effect of the stimuli (*correct_voicing*) is a significant improvement (p-value of \~0.004) on its prediction.

For random effects it seems that the odds of a participant choosing *voiced* varied with the speakers voice (*voice*) with a standard deviation of \~0.903 when the stimulus is *voiced*. This means that some speakers voices made participants lean more toward *voiced* and others toward *voiceless*. The effect of the stimulus was also affected differently for each speakers voice (standard deviation of \~0.451), indicating that some speakers voices changed participants odds of choosing *voiced* more than others when the stimulus went from *voiced* to *voiceless*.

The odds of choosing *voiced* also varied for the pairs of words (*word pair*) when the stimulus was *voiced* (standard deviation of \~0.790), indicating that some pairs made participants lean more toward *voiced* and some more toward *voiceless*. The slope with stimulus also varied here (standard deviation of \~ 0.495), indicating that some pairs made participants more likely to choose *voiced* when the stimulus went from *voiced* to *voiceless*.

The effect of the participant (*listener*) was comparatively much smaller. The intercept had a standard deviation of \~0.191 when the stimulus was *voiced*, indicating that the difference in baseline was small. Some difference (standard deviation of \~0.183) was also seen when the stimulus went from *voiced* to *voiceless*, but, again, small in comparison.

In conclusion we can see that the variability was larger for elements related to the stimulus, and less varied by participants. Participants were more likely to choose *voiced* when the stimulus was *voiceless*.

### (C) Plot predictions

Plot the model predictions alongside a measure of uncertainty (standard error or 95% CI).

Estimating variance components for linear mixed effects models is mathematically not as straightforward as it is for simple linear models. We shortcut manual extraction for now and use the beautiful `ggeffects` package. Check their `predict_response()` function and use it to derive predictions and plot them.

```{r}
#| label: plot_prediction_in

#Generating predictions for *correct_voicing* with *predict_response()*
predictions <- predict_response(
  xmdl,
  terms = "correct_voicing"
)

# Making plot using the predictions
plot_predictions <- 
  predictions %>% 
  ggplot(
    aes(
      x = x,
      y = predicted
    )
  )+
  geom_point()+
  geom_errorbar(
    aes(
      ymin = conf.low, 
      ymax = conf.high
    )
  )+
  labs(
    x = "Correct Voice",
    y = "Predicted likelyhood of choosing *voiced*",
    title = "Predictions (+ 95% CI)"
  )+
  theme_minimal()


# Displaying plot
plot_predictions

```

### (D) OPTIONAL Plot variation across grouping variables

Compare the descriptive averages for the critical relationship for your grouping variables and describe what's going on.

```{r}
#| label: plot_descriptive_in

# YOUR CODE HERE

```

We can see massive variation, particularly for different item pairs and different speaker voices.

### (E) Use Github to upload your analysis

-   fork https://github.com/timo-b-roettger/ling4136-advanced-stats
-   clone your forked repo onto your computer
-   store your solutions in this script as `YOURLASTNAME_04_homework.qmd`
-   move your script into `/homework/04_homework/`
-   commit and push changes to your forked repo
-   add me as a collaborator to your repository and send the link to your homework through Canvas

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
